### 2022-10-08 AISTATS2023/paper.tex

\begin{abstract}
  Accumulated Local Effects (ALE) is a popular approach to explainable
  AI that describes how a feature influences the decisions of a model
  on average, taking into account potential correlations between the
  features.  However, ALE does not provide an explicit component for
  modeling the uncertainty of the effect, i.e., the level of
  heterogeneity of the effects behind the average explanation, which
  is crucial for a complete interpretation when the black-box function
  contains interactions between the correlated features. In this work,
  we define Uncertainty-aware ALE (UALE), an extension of ALE for
  quantifying the heterogeneous effects and we propose a method for
  estimating it. We formally prove the conditions the bin-splitting,
  i.e.~the partitioning the domain of the feature of interest into
  consecutive non-overlapping intervals, must respect for an an
  unbiased estimation of the uncertainty. We finally propose a
  computationally-grounded algorithm for finding the optimal sequence
  of bins given the limited instances of the training set. We
  demonstrate through synthetic and real datasets, that UALE
  quantifies the average and heterogeneous effects correctly and
  approximates them robustly by optimizing the bin-splitting
  procedure.
\end{abstract}

@@

Recently, Machine Learning (ML) has been adopted in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), the field that provides interpretations about the behavior of
complex black-box models. XAI literature distinguishes between local
and global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain the prediction of a
specific instance, whereas global methods provide an explanation that summarizes the entire model behavior. Global methods provide a universal explanation, aggregating the various local explanations into a single interpretable outcome,
usually a number or a plot. If users want to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
is vulnerable to misinterpretations. Under strong interactions and
correlations between features, the global explanation may obfuscate
heterogeneous effects that exist under the
hood~\citep{herbinger2022repid}; a phenomenon called aggregation
bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category of global
explainability methods. The objective of FE is to isolate the impact
of a single feature on the output.~\footnote{Often, FE methods also
  isolate the combined effect of a pair of features to the
  output. Combinations of more than two features are uncommon, since they are difficult to estimate and visualize.}
FE methods suffer from aggregation bias because, often, implications of the average effect to the model behavior are unclear. For example, a feature with zero
average effect may indicate no effect on the output or, in contrast,
highly positive in some cases and highly negative in others.

There are
three widely-used FE methods; Partial Dependence Plots
(PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML cases. Therefore,
ALE has been established as the state-of-the-art FE method. However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one ALE approximation. Regarding the
definition ALE does not inform the user about the uncertainty, i.e.,
the level of heterogeneous effects hidden behind the average effect
due to implicit feature interactions. In contrast, in PDPs the
heterogeneous effects can be identified by exploring the
Individual Conditional Expectations
(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
i.e.~the estimation using the (limited) samples of the training set,
ALE requires the additional \textit{bin-splitting} step. Bin-splitting
consists of partitioning the domain of the feature of interest into consecutive non-overlapping intervals (regions) and estimating a
constant effect in each region. The effect is estimated from the population of samples that fall
inside the region. Specifying an appropriate set of bins is of particular
importance, since ALE's interpretation is meaningful only inside each
region \citep{molnar2022}. However, this crucial step has not raised
the appropriate attention and the approximation is vulnerable to
potential misinterpretations.

>>>>

\begin{abstract}
  Accumulated Local Effects (ALE) is a popular approach to explainable
  AI (XAI) that describes how a feature influences the decisions of a model
  on average, taking into account potential correlations between the
  features.  However, ALE does not provide an explicit component for
  modeling the uncertainty of the effect, i.e., the level of
  heterogeneity of the effects behind the average explanation, which
  is crucial for a complete interpretation when the black-box function
  contains interactions between the correlated features. In this work,
  we define Uncertainty-aware ALE (UALE), an extension of ALE for
  quantifying the heterogeneous effects and we propose a method for
  estimating it. We formally prove the conditions the bin-splitting,
  i.e.~the partitioning the domain of the feature of interest into
  consecutive non-overlapping intervals, must respect for an an
  unbiased estimation of the uncertainty. We finally propose a
  computationally-grounded algorithm for finding the optimal sequence
  of bins given the limited instances of the training set. We
  demonstrate through synthetic and real datasets, that UALE
  quantifies the average and heterogeneous effects correctly and
  approximates them robustly by optimizing the bin-splitting
  procedure.
\end{abstract}

@@

Recently, Machine Learning (ML) has been adopted in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), to understand the decision mechanism of 
%the field that provides interpretations about the behavior of
complex black-box models. XAI literature distinguishes between local
and global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain the prediction of a
specific instance, whereas global methods provide an explanation that summarizes the entire model behavior. Global methods provide a universal explanation, aggregating the various local explanations into a single interpretable outcome,
usually a number or a plot. If users want to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
is vulnerable to misinterpretations. Under strong interactions and
correlations between features, the global explanation may obfuscate
heterogeneous effects that exist under the
hood~\citep{herbinger2022repid}; a phenomenon called \emph{aggregation
bias}~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category of global
explainability methods aiming at isolating the impact
of a single feature on the output\footnote{Often, FE methods also
  isolate the combined effect of a pair of features to the
  output. Combinations of more than two features are uncommon, since they are difficult to estimate and visualize.}.
FE methods suffer from aggregation bias because, often, implications of the average effect on the model behavior are unclear. For example, a feature with zero average effect may indicate no effect on the output or, 
a highly positive effect in some cases and a highly negative effect in others. 
There are
three widely-used FE methods: \emph{Partial Dependence Plots}
(PDP)\citep{friedman2001greedy}, \emph{Marignal Plots}
(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized \todo{Add a reference}for
computing erroneous effects when the input features are (highly)
correlated, which is typical in many ML cases. Therefore,
ALE has been established as the state-of-the-art FE method. However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one ALE approximation. Regarding the
definition, ALE does not inform the user about the uncertainty, i.e.,
the level of heterogeneous effects hidden behind the average effect
due to implicit feature interactions. In contrast, in PDPs the
heterogeneous effects can be identified by exploring the
Individual Conditional Expectations
(ICE)~\citep{goldstein2015peeking}. Regarding the approximation,
i.e.~the estimation using the (limited) samples of the training set,
ALE requires an additional \textit{bin-splitting} step to partition the domain of the feature of interest into consecutive non-overlapping intervals (regions) and to estimate a
constant effect in each region. The effect is estimated from the population of samples that fall
inside the region. Specifying an appropriate set of bins is of particular
importance, since ALE's interpretation is meaningful only inside each
region~\citep{molnar2022}. However, this crucial step has not raised
the appropriate attention and the approximation is vulnerable to
potential misinterpretations.\todo{I miss what exactly is the simplication of existing mehtods here. Should we add e.g., typically equisized bins?}
